Stochastic Gradient Descent

### 2018-04-08 sgd/sgd.tex
updated prez

\begin{frame}
	\frametitle{Optimization is everywhere, and sometimes is easy}
	Many statistical procedures involve minimizing or maximizing some function 
	applied to data \\~\\

	In \textbf{parametric} statistics, we often make assumptions that make this optimization ``nice''
	\begin{itemize}
		\item \small Example: in OLS, we do not need to try
			different values
			of $\hat{\beta}$ to see which minimizes the loss
			function, we (typically) can just evaluate
			$(X'X)^{-1}X'Y$ \\~\\
	\end{itemize}
	\smallskip

	\centering
	\begin{tikzpicture}
		\begin{axis}[
			yticklabels={,,},
			xticklabels={,,},
			width=0.5\textwidth,
			height=0.4\textwidth,
			legend style = {font=\small}
		]
			\addlegendentry{$f=\sum (\hat{y} - y)^2$}
			\addplot[mark=none, color=blue]{x^2};
		\end{axis}
	\end{tikzpicture}
\end{frame}

\begin{frame}
	\frametitle{Other times, optimization is not so easy}
	Suppose that we have a typical supervised classification problem
	\begin{itemize}
		\item \small Non-parametric: no assumptions about distribution of data
		\item Feature vector $\mathbf{X}_i$, label $Y_i$
		\item Want to find best prediction function $f_w^*$ from class $\mathcal{F}$
		\item Optimization: pick weights $\bm{w}$ that minimize
			empirical risk according to some convex
			loss function $L(f_w(x), y)$ \\~\\
	\end{itemize}

	Our lack of assumptions requires a different approach to optimization
	\begin{itemize}
		\item \small Cannot analytically identify stationary point
		\item Need to numerically search for it
	\end{itemize}
\end{frame}

@@

	But, big difference between:
	\begin{itemize}
		\item Speed $:=$ number of iterations
		\item Speed $:=$ time (clock on wall)
	\end{itemize}
\end{frame}

@@

	This becomes computationally intractable as $n$ grows large \\~\\

	Knowing that the quality of our approximation gets linearly or
	quadratically better with $t$ is not comforting if each $t$ takes days to run
\end{frame}

@@

	For each step, gradient is computed for a \textbf{single} randomly
	chosen observation $i$:

	$$
	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w L(f_w(\mathbf{X}_i), y_i)
	$$ \\~\\

@@

\begin{frame}
	\frametitle{SGD is particularly useful when $n$ is large and computation
	time is important}
	\begin{table}[t]
		\begin{tabular}{|l|c|c|}
			\hline
			& \textbf{GD} & \textbf{SGD} \\
			\textbf{Time to accuracy} $\mathbf{\rho}$ & $n \log
			\frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
			\hline
		\end{tabular}
	\end{table}

@@

	...SGD probably is involved. Example: Google's \textbf{AlphaGo} program.

	\begin{figure}[b]
	\centering
	\includegraphics[scale=0.25]{go}
	\end{figure}

>>>>

\begin{frame}
	\frametitle{Optimization is fundamental to statistical modeling}
	Suppose that we have a typical supervised classification problem
	\begin{itemize}
		\item \small Non-parametric: no assumptions about distribution of data
		\item Feature vector $\mathbf{X}_i$, label $Y_i$
		\item Want to find best prediction function $f_w^*$ from class $\mathcal{F}$
		\item Optimization: pick weights $\hat{\bm{w}}$ that minimize
			empirical risk according to some convex
			loss function $L(f_w(x), y)$ \\~\\
	\end{itemize}

	Our lack of assumptions takes away some familiar tools for finding $\hat{\mathbf{w}}$
	\begin{itemize}
		\item \small Cannot analytically identify $\hat{\mathbf{w}}$
			(e.g. in OLS $=(X'X)^{-1}X'Y$))
		\item \small Newton-Raphson requires 2nd derivative of $L$,
			which might be a pain (unlike in, for example, GLM)\\~\\
	\end{itemize}

	But we still know that if $L$ is convex, there is a unique global
	minimum, and the gradient at that point must be 0
\end{frame}

@@

@@

	This becomes computationally intractable as $n$ grows large 
	\begin{itemize}
		\item \small If we have $n$ = 10 million, we have to evaluate the
			gradient 10 million times for every step (and the
			algorithm may take many steps to converge)\\~\\
		\item \small May take hours or days to converge \\~\\
	\end{itemize}
\end{frame}

@@

	For each step, gradient is computed for a \textcolor{red}{single} randomly
	chosen observation $i$:

	$$
	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w
	L(f_w(\mathbf{X}_{\textcolor{red}{i}}), y_{\textcolor{red}{i})}
	$$ \\~\\

@@

\begin{frame}
\frametitle{SGD is noisier than batch GD}
	\centering
	\includegraphics[scale=0.3]{noise}

	\tiny Source: towardsdatascience.com
\end{frame}

\begin{frame}
	\frametitle{SGD is useful when $n$ is large \& compute
	time is important}
	\begin{table}[t]
		\begin{tabular}{|l|c|c|}
			\hline
			& \textbf{GD} & \textbf{SGD} \\
			\textbf{Time to accuracy} $\mathbf{\rho}$ & $n \log
			\frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
			\hline
		\end{tabular}
	\end{table}

@@

	...SGD probably is involved. \\~\\

	Example: Google's \textbf{AlphaGo} program.
\end{frame}

\begin{frame}
	\begin{figure}[b]
	\centering
	\includegraphics[scale=0.4]{go}
	\end{figure}
\end{frame}

\begin{frame}
	\frametitle{Google slides from ICML 2016}
	\centering
	\includegraphics[scale=0.3]{screenshot}
\end{frame}
